Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-30 RevisedSubmission/hines_hetland_2017_gji.tex
did more

We consider three potential choices for $T_i$. The first option is the
one-dimensional SE covariance function,
\begin{equation}\label{eq:TimeSE}
T_i(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right).
\end{equation}
Note that $T_i$ includes the parameter $\phi$, which serves to scale
the covariance function $C_{u_i}$. The second option is a member of
the Wendland class of covariance functions \citep{Wendland2005}.
Wendland covariance functions have compact support and hence their
corresponding covariance matrices are sparse. In our analysis, we
exploit this sparsity with the CHOLMOD software package
\citep{Chen2008}. Wendland covariance functions can be constructed
such that they are positive definite in $\mathbb{R}^d$ and their
corresponding Gaussian process can be differentiated $k$ times. We use
$d=1$, because we are describing the temporal covariance of $u_i$, and
we use $k=2$, giving samples of $u_i$ continuous velocities and
accelerations. The corresponding Wendland covariance function is
\begin{equation}\label{eq:Wendland} 
T_i(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\tau}\right)^5_+ 
            \left(\frac{8|t - t'|^2}{\tau^2} + \frac{5|t - t'|}{\tau} + 1\right), 
\end{equation} 
where $(t)_+ = \mathrm{max}(0,t)$. The third option for $T_i$ is the
covariance function for integrated Brownian motion (IBM). IBM is a
Gaussian process with zero mean and its covariance function can be
found by integrating the covariance function for Brownian motion,
\begin{align}\label{eq:IBM}
T_i(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(s,s') \,ds'\,ds \\
          &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used in the context of Kalman filtering as a
non-parametric model for the time dependence of geophysical signals
\citep[e.g.,][]{Segall1997, McGuire2003, Ohtani2010, Hines2016a}.
Similar to Brownian motion, IBM has a reference time, $t=0$, at which
the process begins. For some geophysical signals, it is appropriate to
have this reference time. For example, if we are trying to identify
postseismic deformation then $t$ should be zero at the time of the
earthquake.  However, if we are interesting in detecting transient
events, where there is no known start time, then IBM may not be an
appropriate prior model. Instead, one may prefer to use the SE or
Wendland covariance functions because they are stationary. In the
following analysis, we make the choice that $t$ is zero on the first
epoch of $\data_i^*$. Using an earlier reference time does not change
the results discussed in this section.

@@

We use the REML method to determine the appropriate values for the
parameters in $X_i$ and $T_i$, $\mitbf{\theta} = \{\ell, \phi, \tau
\}$. we divide the GNSS data into seven subsets that are four months
long and each centered on the time of a SSE. The times of the seven
SSEs are determined with tremor records from \cite{Wech2010}. We find
the optimal $\mitbf{\theta}$ for each subset of data, for each
displacement component, and for each choice of $T_i$. We choose to
make each data subsets four months long because it is long enough to
encompass a SSE in the Pacific Northwest, while it is still
computationally tractable to estimate $\mitbf{\theta}$. However, four
months is too short to resolve the sinusoids in $\data_i$, and they
are omitted from $\data_i$ in this REML analysis. The REML likelihood
function that we are maximizing with respect to $\mitbf{\theta}$ is
now 
\begin{equation}\label{eq:REMLPrior} 
\mathcal{L}(\mitbf{\theta}) =
\left(\frac{\left| \G^T \G \right|}{(2\pi)^{n-6m}
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2} \data_i^{*T} \mitbf{K}(\mitbf{\theta}) \data_i^*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G
                            \left( \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right)^{-1}
                            \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1}
\end{equation}
and $\mitbf{\Sigma}(\mitbf{\theta}) = \mitbf{C}_{\eta_i} +
C_{u_i}(\points,\points;\mitbf{\theta})$. In the above equation,
$\data^*_i$ consists of a single subset of data. The estimated
parameters are summarized in Table 1. Based on the interquartile
ranges, the estimated parameters for the SE and Wendland covariance
functions do not vary significantly between SSEs. This suggests that
the median of estimated parameters should be an appropriate prior
model for all Pacific Northwest SSEs. For the IBM model, there are
several anomalously large values for $\ell$ and $\phi$, which is why
the interquartile ranges are large.

\begin{table}\label{tab:Parameters}
\caption{
Optimal parameters for the prior on transient displacements determined
with the REML method. The temporal covariance function is indicated by
the ``$T$'' column. The SE, IBM, and Wendland covariance functions are
defined in eqs. (\ref{eq:TimeSE}), (\ref{eq:IBM}), and
(\ref{eq:Wendland}), respectively. The spatial covariance function,
$X$, is the squared exponential (eq. \ref{eq:SE}) in all cases. The
parameters are estimated for each of the seven SSEs considered in
this study, and the tabulated values indicate the median and
interquartile ranges of estimates. The ``diff $\log$(REML)'' column
compares the log REML likelihood to the log REML likelihood when using
the SE covariance function for $T$. Negative values indicate that
observations are more consistent with the SE covariance function.
} 
\begin{tabular} {l l l l l l}
$T$ & direction & $\ell$  & $\phi$   & $\tau$  & diff. $\log$(REML) \\ \hline
SE & east   & 92 $\pm$ 25 km  & 0.62 $\pm$ 0.11 mm  & 0.026 $\pm$ 0.011 yr  &  - \\
SE & north  & 91 $\pm$ 53 km  & 0.43 $\pm$ 0.05 mm  & 0.030 $\pm$ 0.017 yr  &  - \\
Wendland & east   & 95 $\pm$ 30 km  & 0.66 $\pm$ 0.15 mm  & 0.093 $\pm$ 0.044 yr &  0.78 $\pm$ 0.87 \\
Wendland & north  & 92 $\pm$ 57 km  & 0.46 $\pm$ 0.10 mm  & 0.116 $\pm$ 0.057 yr &  0.08 $\pm$ 0.58 \\
IBM & east   & 110 $\pm$ 130 km & 290 $\pm$ 420 mm/yr$^{1.5}$  & -          & -16.4 $\pm$ 7.8 \\
IBM & north  & 150 $\pm$ 560 km & 110 $\pm$ 250 mm/yr$^{1.5}$ & -           & -10.1 $\pm$ 2.3 \\
\end{tabular}
\end{table}

Next we identify which covariance function for $T$ best describes the
SSEs. One approach is to compare the REML likelihoods for each
covariance function, similar to the analysis in \citet{Langbein2004}.
In Table 1, we summarize how the log REML likelihoods for the Wendland
and IBM covariance functions compare to the SE covariance function.
Based on the differences in log REML likelihoods, the data is
substantially more likely to come from a Gaussian process with a SE or
Wendland covariance function than an IBM covariance function. The REML
likelihoods do not definitively indicate whether the SE or Wendland
covariance function is preferable.

To further explore which covariance function for $T$ best describes
SSEs, we compare the observations to the predicted displacements for
each covariance function. We consider the data prediction vector to be
$\hat{\data} = \left(u(\points) + \G\mitbf{a}\right)|\data_*$.
Following a similar procedure as in Section \ref{sec:Method}, it can
be shown that $\hat{\data}$ is normally distributed with mean
\begin{equation}\label{eq:DataPredMean}
\mitbf{\mu}_{\hat{d}} = \left[\begin{array}{cc}
                           C_u(\points,\points) & \G \\
                           \end{array}\right]
                     \left[\begin{array}{cc}
                           \mitbf{\Sigma} & \G \\
                           \G^T  & \zeros \\
                           \end{array}\right]^{-1}
                     \left[\begin{array}{c}
                           \data_* \\
                           \zeros \\
                           \end{array}\right]
\end{equation}  
and covariance
\begin{equation}\label{eq:DataPredCov}
\mitbf{C}_{\hat{\data}} = C_u(\points,\points) - 
                        \left[\begin{array}{cc}
                              C_u(\points,\points) & \G \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \mitbf{\Sigma} & \G \\
                              \G^T  & \zeros \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_u(\points,\points) \\
                              \G^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\data}$ using SE, Wendland, and IBM covariance
functions for $T$ and the median parameters from Table 1. Figure
\ref{fig:Fit} compares the easting component of $\data_*$ to
$\hat{\data}$ during the winter 2015-2016 SSE at the three stations
that record the strongest SSE signal, ALBH, P436, and SC02. Based on
Figure \ref{fig:Fit}, $\hat{\data}$ is insensitive to the choice of
covariance function for $T$. The predicted displacements for the IBM
covariance function contain slightly more high frequency, and perhaps
spurious, features. When comparing $\hat{\data}$ to $\data_*$ at
station ALBH and SC02, it appears that $\hat{\data}$ slightly
underestimates the rate of deformation during the SSE, regardless of
the chosen covariance function. This over-smoothing could indicate
that the chosen time-scale parameter, $\tau$, is too large. The
predicted displacements at the remaining stations, which record more
subtle SSE deformation, seem to faithfully describe $\data_*$.

For our estimates of transient strain discussed in the next section,
we ultimately settle on the Wendland covariance function for $T$ and
use the median values from Table 1 for the parameters. We choose
the Wendland covariance function over the SE covariance function
because of its computational advantages.

\begin{figure*}
\includegraphics{figures/signal_fit/signal-fit.pdf}
\caption{
Easting component of the observed displacements, $\data_*$, and
predicted displacements, $\hat{\data}$, during an SSE at three
stations where the SSE signal is strongest. The predicted
displacements are shown for when $T$ is a squared exponential (SE),
Wendland, and integrated Brownian motion (IBM) covariance function.
The one standard deviation uncertainties are shown for the
observations and the SE predictions. For clarity, uncertainties are
not shown for the IBM and Wendland predictions, but they are nearly
equivalent to the uncertainties for the SE predictions. The SE and
Wendland predictions are practically indistinguishable.
}   
\label{fig:Fit}
\end{figure*}

>>>>

We consider three potential choices for $T_i$. The first option is the
one-dimensional SE covariance function,
\begin{equation}\label{eq:TimeSE}
T_i(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right).
\end{equation}
Note that $T_i$ includes the parameter $\phi$, which serves to scale
the covariance function $C_{u_i}$. The second option is a member of
the Wendland class of covariance functions \citep{Wendland2005}.
Wendland covariance functions have compact support and hence their
corresponding covariance matrices are sparse. In our analysis, we
exploit this sparsity with the CHOLMOD software package
\citep{Chen2008}. Wendland covariance functions can be constructed
such that they are positive definite in $\mathbb{R}^d$ and their
corresponding Gaussian process can be differentiated $k$ times. We use
$d=1$, because we are describing the temporal covariance of $u_i$, and
we use $k=2$, giving samples of $u_i$ continuous velocities and
accelerations. The corresponding Wendland covariance function is
\begin{equation}\label{eq:Wendland} 
T_i(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\tau}\right)^5_+ 
            \left(\frac{8|t - t'|^2}{\tau^2} + \frac{5|t - t'|}{\tau} + 1\right), 
\end{equation} 
where $(t)_+ = \mathrm{max}(0,t)$. The third option for $T_i$ is the
covariance function for integrated Brownian motion (IBM). IBM is a
Gaussian process with zero mean and its covariance function can be
found by integrating the covariance function for Brownian motion,
\begin{align}\label{eq:IBM}
T_i(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(s,s') \,ds'\,ds \\
          &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used in the context of Kalman filtering as a
non-parametric model for the time dependence of geophysical signals
\citep[e.g.,][]{Segall1997, McGuire2003, Ohtani2010, Hines2016a}.
Similar to Brownian motion, IBM has a reference time, $t=0$, at which
the process begins. For some geophysical signals, it is appropriate to
have this reference time. For example, if we are trying to identify
postseismic deformation then $t$ should be zero at the time of the
earthquake.  However, if we are interested in detecting transient
events, where there is no known start time, then IBM may not be an
appropriate prior model. Instead, one may prefer to use the SE or
Wendland covariance functions because they are stationary. In the
following analysis, we make the choice that $t$ is zero on the first
epoch of $\data_i^*$. Using an earlier reference time does not change
the results discussed in this section.

@@

We use the REML method to determine the appropriate values for the
parameters in $X_i$ and $T_i$, $\mitbf{\theta} = \{\ell, \phi, \tau
\}$. We divide the GNSS data into seven subsets that are four months
long and each centered on the time of a SSE. The times of the seven
SSEs are determined with tremor records from \cite{Wech2010}. We
choose to make each data subsets four months long because it is long
enough to encompass a SSE in the Pacific Northwest, while it is still
computationally tractable to estimate $\mitbf{\theta}$. However, four
months is too short to resolve the sinusoids in $\data_i$, and they
are omitted from $\data_i$ in this REML analysis. We find the optimal
$\mitbf{\theta}$ for each subset of data, for each displacement
component, and for each choice of $T_i$. The REML likelihood function
that we are maximizing with respect to $\mitbf{\theta}$ is now
\begin{equation}\label{eq:REMLPrior} 
\mathcal{L}(\mitbf{\theta}) =
\left(\frac{\left| \G^T \G \right|}{(2\pi)^{n-6m}
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2} \data_i^{*T} \mitbf{K}(\mitbf{\theta}) \data_i^*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G
                            \left( \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right)^{-1}
                            \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1}
\end{equation}
and $\mitbf{\Sigma}(\mitbf{\theta}) = \mitbf{C}_{\eta_i} +
C_{u_i}(\points,\points;\mitbf{\theta})$. In the above equation,
$\data^*_i$ consists of a single subset of data. The estimated
parameters are summarized in Table 1. Based on the interquartile
ranges, the estimated parameters for the SE and Wendland covariance
functions do not vary significantly between SSEs. This suggests that
the median of the estimated parameters should be appropriate for all
the Pacific Northwest SSEs. For the IBM model, there are several
anomalously large values for $\ell$ and $\phi$, which is why the
interquartile ranges are large.

\begin{table}\label{tab:Parameters}
\caption{
Optimal values for the parameters in $X_i$ and $T_i$ determined with
the REML method. The covariance function for $T_i$ is indicated by the
``$T_i$'' column. The SE, Wendland, and IBM covariance functions are
defined in eqs. (\ref{eq:TimeSE}), (\ref{eq:Wendland}), and
(\ref{eq:IBM}), respectively. We use eq. (\ref{eq:SE}) for $X_i$.
The parameters are estimated for each of the seven SSEs considered in
this study, and the tabulated values indicate the median and
interquartile ranges of estimates. The ``diff $\log$(REML)'' column
compares the optimal log REML likelihoods to those for when $T_i$ is
the SE covariance function. Negative values indicate that observations
are more consistent with the SE covariance function.
} 
\begin{tabular} {l l l l l l}
$T_i$ & component & $\ell$  & $\phi$   & $\tau$  & diff. $\log$(REML) \\ \hline
SE & east   & 92 $\pm$ 25 km  & 0.62 $\pm$ 0.11 mm  & 0.026 $\pm$ 0.011 yr  &  - \\
SE & north  & 91 $\pm$ 53 km  & 0.43 $\pm$ 0.05 mm  & 0.030 $\pm$ 0.017 yr  &  - \\
Wendland & east   & 95 $\pm$ 30 km  & 0.66 $\pm$ 0.15 mm  & 0.093 $\pm$ 0.044 yr &  0.78 $\pm$ 0.87 \\
Wendland & north  & 92 $\pm$ 57 km  & 0.46 $\pm$ 0.10 mm  & 0.116 $\pm$ 0.057 yr &  0.08 $\pm$ 0.58 \\
IBM & east   & 110 $\pm$ 130 km & 290 $\pm$ 420 mm/yr$^{1.5}$  & -          & -16.4 $\pm$ 7.8 \\
IBM & north  & 150 $\pm$ 560 km & 110 $\pm$ 250 mm/yr$^{1.5}$ & -           & -10.1 $\pm$ 2.3 \\
\end{tabular}
\end{table}

% Choosing the covariance function for T_i with REML

Next, we identify which covariance function for $T_i$ best describes
the time dependence of deformation from SSEs. One approach is to
choose the covariance function that produces the largest optimal REML
likelihoods, similar to the analysis in \citet{Langbein2004}. In Table
1, we summarize how the optimal REML likelihoods for the Wendland and
IBM covariance functions compare to those for the SE covariance
function. Based on the differences in optimal REML likelihoods, the
data is substantially more likely to come from a Gaussian process with
a SE or Wendland covariance function than an IBM covariance function.
The REML likelihoods do not definitively indicate whether the SE or
Wendland covariance function is preferable.

% Choosing the covariance function for T_i with posterior fit

A more intuitive way of deciding which function to use for $T_i$ is to
compare the posterior displacements and the observed displacements.
The posterior displacements consist of our estimate of transient
displacements, secular trends, and seasonal deformation. Specifically,
the posterior displacement vector is $\hat{\data}_i = \left(
u_i(\points) + \G\mitbf{m}_i \right) | \left( \data_i = \data_i^*
\right)$. Following a similar procedure as in Section
\ref{sec:Method}, it can be shown that $\hat{\data}_i$ has a
multivariate normal distribution with mean vector
\begin{equation}\label{eq:DataPredMean}
\mitbf{\mu}_{\hat{d}_i} = \left[\begin{array}{cc}
                          C_{u_i}(\points,\points) & \G \\
                          \end{array}\right]
                          \left[\begin{array}{cc}
                          \mitbf{C}_{\eta_i} + C_{u_i}(\points,\points) & \G \\
                          \G^T  & \zeros \\
                          \end{array}\right]^{-1}
                          \left[\begin{array}{c}
                          \data_i^* \\
                          \zeros \\
                          \end{array}\right]
\end{equation}  
and covariance matrix
\begin{equation}\label{eq:DataPredCov}
\mitbf{C}_{\hat{d}_i} = C_{u_i}(\points,\points) - 
                        \left[\begin{array}{cc}
                              C_{u_i}(\points,\points) & \G \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \mitbf{C}_{\eta_i} + C_{u_i}(\points,\points) & \G \\
                              \G^T  & \zeros \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_{u_i}(\points,\points) \\
                              \G^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\data}_i$ using the SE, Wendland, and IBM covariance
functions for $T_i$, and we use the median values from Table 1 for
$\mitbf{\theta}$. Figure \ref{fig:Fit} compares $\data_i^*$ to
$\hat{\data}_i$ for the winter 2015-2016 SSE at stations ALBH, P436,
and SC02. Based on Figure \ref{fig:Fit}, the chosen covariance
function for $T_i$ has almost no effect on $\hat{\data}_i$. The
posterior displacements for the IBM covariance function contain
slightly more high frequency, and perhaps spurious, features.
Regardless of the chosen covariance function, $\hat{\data}_i$ appears
to underestimate the rate of deformation during the SSE at stations
ALBH and SC02. The deformation at these two stations is particularly
rapid compared to the surrounding stations, and the misfit between
$\data_i^*$ and $\hat{\data}_i$ is likely due to over-smoothing. This
over-smoothing could indicate that the chosen values for $\tau$ or
$\ell$ are too large. However, $\hat{\data}_i$ does faithfully fit
$\data_i^*$ at the remaining stations, and so we do not attempt to
further refine the parameters.

% The choice

For our estimates of transient strain discussed in the next section,
we ultimately settle on the Wendland covariance function for $T_i$,
and we use the median values from Table 1 for the parameters. We
choose the Wendland covariance function over the SE covariance
function because of its computational advantages.

\begin{figure*}
\includegraphics{figures/signal_fit/signal-fit.pdf}
\caption{
Easting component of the observed displacements, $\data_\e^*$, and
posterior displacements, $\hat{\data}_\e$, during the winter 2015-2016
SSE. We show $\hat{\data}_\e$ for when $T_i$ is a SE, Wendland, and
IBM covariance function. The one standard deviation uncertainties are
shown for $\data_\e^*$ and the $\hat{\data}_\e$ that has a SE
covariance function for $T_i$. For clarity, uncertainties are not
shown for $\hat{\data}_\e$ when $T_i$ is an IBM or Wendland covariance
function, but the uncertainties are nearly equivalent to the
uncertainties for when $T_i$ is the SE covariance function. The
posterior displacements for the different covariance functions are all
practically indistinguishable.
}   
\label{fig:Fit}
\end{figure*}
